### 2020-08-20 submission/manuscript.Rmd
Additional edits

## Abstract
The gut microbiota has a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single source. We treated mice from 6 sources (2 University of Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by *C. difficile* challenge 1 day later and then measured *C. difficile* colonization levels through 9 days post-infection. The microbiota were profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all mice were colonized 1-day post-infection, variation emerged from days 3-7 post-infection with animals from some sources colonized with *C. difficile* for longer and at higher levels. We identified bacteria that varied in relative abundance across sources and throughout the experiment. Some bacteria were consistently impacted by clindamycin treatment in all sources of mice including *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae*. To identify bacteria that were most important to colonization regardless of the source, we created logistic regression models that successfully classified mice based on whether they cleared *C. difficile* by 7 days post-infection using baseline, post-clindamycin, and 1-day post-infection community composition data. With these models, we identified 4 bacteria that were predictive of whether *C. difficile* cleared. They varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*). Microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.

@@

Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs) due to there effect on the intestinal microbiota, but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012]. Additionally, asymptomatic colonization, where *C. difficile* is detectable, but symptoms are absent, has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018]. The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020]. However, it is not clear how much inter-individual microbiota variation contributes to the range of outcomes observed after *C. difficile* exposure relative to other risk factors.

Mouse models of CDIs have been a great tool for understanding *C. difficile* pathogenesis [@Hutton2014]. The number of CDI mouse model studies has grown substantially since Chen et al. published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C. difficile* colonization and symptoms such as diarrhea and weight loss [@Chen2008]. CDI mouse models have been used to examine translationally relevant questions regarding *C. difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012]. However, variation in the microbiota between mice from the same breeding colony is much less than the inter-individual variation observed between humans [@Baxter2014; @Nagpal2018]. Studying CDIs in mice with a homogenous microbiota is likely to overstate the importance of individual mechanisms. Using mice that have a more heterogenous microbiota would allow researchers to identify and validate more generalizable mechanisms responsible for CDI.

In the past, our group has attempted to introduce more variation into the mouse microbiota by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018]. An alternative approach to maximize microbiota variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020]. The differences between the microbiota of mice from vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. Different research groups have also observed variation in CDI outcomes despite using similar murine models [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014]. Here we examined how variations in the baseline microbiota and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C. difficile* colonization and the time needed to clear the infection.

@@

**The variation in the microbiota is high between mice from different sources.** We obtained C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other in 2010 (the Young and Schloss lab colonies) and four commercial sources: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they are commonly used for murine CDI studies [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. Two experiments were conducted, approximately 3 months apart.

We sequenced the 16S rRNA gene from fecal samples collected from these mice after they acclimated to the University of Michigan animal housing environment. We first examined the alpha diversity across the 6 sources of mice. There was a significant difference in the richness (i.e. number of observed operational taxonomic units (OTUs)), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across the sources of mice (Fig. 1A-B and Tables S1-2). Next, we compared the community structure of mice (Fig. 1C). The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`, Fig. 1C and Table S3). Mice that are co-housed tend to have similar gut microbiotas due to coprophagy [@Nguyen2015]. Since mice within the same source were housed together, it was not surprising that the cage effect also contributed to the observed community variation. There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of the vendors also clustered by experiment (Fig. S2C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences at the community level, we next identified the bacteria that varied between sources of mice. There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources (Fig. 1D and Table S5). Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (all *P* > 0.05). By using mice from six sources we were able to increase the variation in the starting communities to evaluate in a clindamycin-based CDI model.

@@

#Combine into a data frame and figure out percent colonized per day
percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
  group_by(day) %>% 
  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(percent_colonized)
n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
  pull(total_n)
n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
  pull(total_n)
n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
  pull(total_n)
n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
  pull(total_n)
n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
  pull(total_n)
n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
  pull(total_n)
n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(total_n)
n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
  pull(total_n)
n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
  pull(total_n)
```
**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization, but clearance time varies across sources.** Clindamycin is frequently implicated with human CDIs [@Guh2018] and was part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. We have previously demonstrated mice are rendered susceptible to *C. difficile*, but clear the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019]. All mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 2A). The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig. 2B). However, between 3 and 7 days post-infection, we observed variation in *C. difficile* levels across sources of mice (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 2B and Table S6). This suggested the mouse source was associated with *C. difficile* clearance. While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the 1st experiment compared to the 2nd and the Envigo mice took longer to clear in the 2nd experiment compared to the 1st (Fig. S2A-B). The change in the mice's weight significantly varied across sources of mice with the most weight loss occurring two days post-infection (Fig. 2C and Table S7). There was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice). This was particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C. difficile* (Fig. S2E). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. 2B) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.

```{r clindamycin_taxa}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the variation in fecal communities that we observed across breeding colonies, we hypothesized that variation in *C. difficile* clearance would be explained by community variation across the 6 sources of mice. As expected, clindamycin treatment decreased the richness and Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly, significant differences in diversity metrics between sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Tables S1-2). The clindamycin treatment decreased the variation in community structures between sources of mice. The source of mice and the interactions between source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig. 3C and Table S3). However, there were only `r source_otus_d0` OTUs (Fig. 3D and Table S8) with relative abundances that significantly varied between sources. Next, we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment. We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment in most mice (Fig. 3E and Table S9). When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig. 3D-E and Tables S8-9).  Importantly, some of the OTUs that varied between sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment (Fig. 3D), but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig. 3D). These findings demonstrate that clindamycin had a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary between sources.

**Microbiota variation between sources is maintained after *C. difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Tables S1-2). Although the Charles River mice had more diverse communities and were also able to clear *C. difficile* faster than the other sources, diversity did not explain the observed variation in *C. difficile* colonization across sources. The Young and Schloss mice had the lowest diversity 1 day post-infection and were able to clear *C. difficile* earlier than Jackson, Taconic and Envigo mice. The source of mice and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig. 4C and Table S3). One day after *C. difficile* challenge, there were `r source_otus_d1` OTUs (Fig. 4D and Table S10) with significantly different relative abundances across sources.

Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11). Mice fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment. By 7 days post-infection, when approximately `r format(100-percent_colonized_d7, digits = 1)`% mice had cleared *C. difficile*, most of the mice had not recovered to their baseline community structure (Fig. 4E). The distance to the baseline community did not explain the variation in *C. difficile* clearance as the Schloss and Young mice had mostly cleared *C. difficile*, but their communities were a greater distance from baseline 7 days post-infection compared to the Jackson and Taconic mice that were still colonized. In summary, mouse bacterial communities varied significantly between sources throughout the course of the experiment and a consistent subset of bacteria remained different between sources regardless of clindamycin and *C. difficile* challenge.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
```
**Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C. difficile* by 7 days post-infection.** After identifying taxa that varied between sources, changed after clindamycin treatment, or both, we determined which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic regression models with input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), and post-infection (day = 1) timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12). The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons, Fig. S3C and Table S13). Thus, we were able to use bacterial relative abundance data from the time of *C. difficile* challenge to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile* at that timepoint. This result suggests the bacterial community's response to clindamycin treatment had the greatest influence on subsequent *C. difficile* colonization dynamics.

To examine the bacteria that were driving each model’s performance, we selected the 20 OTUs that had the highest absolute feature weights in each of the 3 models (Table S14). First, we looked at OTUs from the model with the best performance, which was based on the post-clindamycin treatment (day 0) bacterial community data. Out of the 10 highest ranked OTUs, 7 OTUs (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*) were associated with *C. difficile* colonization 7 days post-infection, while 3 OTUs (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*) were associated with clearance (Fig. 5A). Next, we examined whether any of the top 20 ranked OTUs from the post-clindamycin (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data. We identified 6 OTUs (*`r list_overlapping_across_models`*) that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (Table S14). Thus, a subset of bacterial OTUs were important for determining *C. difficile* colonization dynamics across multiple timepoints.

To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we identified the OTUs from each model that varied between sources (Fig. 1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features important to the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 5B). Together, these results suggest that the initial bacterial communities and their responses to clindamycin influence the clearance of *C. difficile*.

>>>>

## Abstract
The gut microbiota has a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use animals obtained from a single source. We treated mice from 6 sources (2 University of Michigan colonies and 4 commercial vendors) with clindamycin, followed by a *C. difficile* challenge and then measured *C. difficile* colonization levels throughout the infection. The microbiota were profiled via 16S rRNA gene sequencing to examine the variation across sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all mice were colonized 1-day post-infection, variation emerged from days 3-7 post-infection with animals from some sources colonized with *C. difficile* for longer and at higher levels. We identified bacteria that varied in relative abundance across sources and throughout the experiment. Some bacteria were consistently impacted by clindamycin treatment in all sources of mice including *Lachnospiraceae*, *Ruminococcaceae*, and *Enterobacteriaceae*. To identify bacteria that were most important to colonization regardless of the source, we created logistic regression models that successfully classified mice based on whether they cleared *C. difficile* by 7 days post-infection using community composition data at baseline, post-clindamycin, and 1-day post-infection. With these models, we identified 4 bacteria that were predictive of whether *C. difficile* cleared. They varied across sources (*Bacteroides*), were altered by clindamycin (*Porphyromonadaceae*), or both (*Enterobacteriaceae* and *Enterococcus*). Allowing for microbiota variation across sources better emulates human inter-individual variation and can help identify bacterial drivers of phenotypic variation in the context of CDIs.

@@

Antibiotics are a common risk factor for *Clostridioides difficile* infections (CDIs) due to their effect on the intestinal microbiota, but there is variation in who goes on to develop severe or recurrent CDIs after exposure [@Teng2019; @Kelly2012]. Additionally, asymptomatic colonization, where *C. difficile* is detectable, but symptoms are absent, has been documented in infants and adults [@Zacharioudakis2015; @Crobach2018]. The intestinal microbiota has been implicated in asymptomatic colonization [@Zhang2015; @VanInsberghe2020], susceptibility to CDIs [@Mancabelli2017, @Duvallet2017], and adverse CDI outcomes [@Seekatz2016; @Khanna2016; @Pakpour2017; @Lee2020]. However, it is not clear how much inter-individual microbiota variation contributes to the range of outcomes observed after *C. difficile* exposure relative to other risk factors.

Mouse models of CDIs have been a great tool for understanding *C. difficile* pathogenesis [@Hutton2014]. The number of CDI mouse model studies has grown substantially since Chen et al. published their C57BL/6 model in 2008, which disrupted the gut microbiota with antibiotics to enable *C. difficile* colonization and symptoms such as diarrhea and weight loss [@Chen2008]. CDI mouse models have been used to examine translationally relevant questions regarding *C. difficile*, including the role of the microbiota and efficacy of potential therapeutics for treating CDIs [@Best2012]. However, variation in the microbiota between mice from the same breeding colony is much less than the inter-individual variation observed between humans [@Baxter2014; @Nagpal2018]. Studying CDIs in mice with a homogeneous microbiota is likely to overstate the importance of individual mechanisms. Using mice that have a more heterogeneous microbiota would allow researchers to identify and validate more generalizable mechanisms responsible for CDI.

In the past, our group has attempted to introduce more variation into the mouse microbiota by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018]. An alternative approach to maximize microbiota variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020]. The differences between the microbiota of mice from vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. Different research groups have also observed different CDI outcomes despite using similar murine models [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014]. Here we examined how variation in the baseline microbiota and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C. difficile* colonization and the time needed to clear the infection.

@@

**The variation in the microbiota is high between mice from different sources.** We obtained C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other in 2010 (the Young and Schloss lab colonies) and four commercial vendors: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they are commonly used for murine CDI studies [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. Two experiments were conducted, approximately 3 months apart.

We sequenced the 16S rRNA gene from fecal samples collected from these mice after they acclimated to the University of Michigan animal housing environment. We first examined the alpha diversity across the 6 sources of mice. There was a significant difference in the richness (i.e. number of observed operational taxonomic units (OTUs)), but not Shannon diversity index across the sources of mice (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively; Fig. 1A-B and Tables S1-2). Next, we compared the community structure of mice (Fig. 1C). The source of mice and the interactions between the source and cage effects explained most of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r dn1_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig. 1C and Table S3). Mice that are co-housed tend to have similar gut microbiotas due to coprophagy [@Nguyen2015]. Since mice within the same source were housed together, it was not surprising that the cage effect also contributed to the observed community variation. There were some differences between the 2 experiments we conducted, as the experiment and cage effects significantly explained the observed community variation for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of the vendors also clustered by experiment (Fig. S2C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences at the community level, we next identified the bacteria that varied between sources of mice. There were `r source_otus_dn1` OTUs with relative abundances that were significantly different between the sources (Fig. 1D and Table S5). Though we saw differences between experiments at the community level, there were no OTUs that were significantly different between experiments within Schloss, Young, and Envigo mice (all *P* > 0.05). By using mice from six sources we were able to increase the variation in the starting communities to evaluate in a clindamycin-based CDI model.

@@

#Combine into a data frame and figure out percent colonized per day
percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
  group_by(day) %>% 
  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(percent_colonized)
n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
  pull(total_n)
n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
  pull(total_n)
n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
  pull(total_n)
n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
  pull(total_n)
n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
  pull(total_n)
n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
  pull(total_n)
n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(total_n)
n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
  pull(total_n)
n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
  pull(total_n)
```
**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization, but clearance time varies across sources.** Clindamycin is frequently implicated with human CDIs [@Guh2018] and was part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. We have previously demonstrated mice are rendered susceptible to *C. difficile*, but clear the pathogen within 9 days when treated with clindamycin alone [@Jenior2018; @Tomkovich2019]. All mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 2A). The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig. 2B). However, between 3 and 7 days post-infection, we observed variation in *C. difficile* levels across sources of mice (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 2B and Table S6). This suggested the source of mice was associated with *C. difficile* clearance. While the colonization dynamics were similar between the two experiments, the Schloss mice took longer to clear in the first experiment compared to the second and the Envigo mice took longer to clear in the second experiment compared to the first (Fig. S2A-B). The change in the mice's weight significantly varied across sources of mice with the most weight loss occurring two days post-infection (Fig. 2C and Table S7). There was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice). This was particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C. difficile* (Fig. S2E). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment  (Fig. 2B). Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.

```{r clindamycin_taxa}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the variation in fecal communities that we observed across breeding colonies, we hypothesized that variation in *C. difficile* clearance would be explained by community variation across the 6 sources of mice. As expected, clindamycin treatment decreased the richness and Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly, significant differences in diversity metrics between sources emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (*P*~FDR~ < 0.05; Fig 3A-B and Tables S1-2). The clindamycin treatment decreased the variation in community structures between sources of mice. The source of mice and the interactions between source and cage effects explained almost all of the observed variation between communities (combined R^2^ = `r d0_rsq_sum`, *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig. 3C and Table S3). However, there were only `r source_otus_d0` OTUs with relative abundances that significantly varied between sources (Fig. 3D and Table S8). Next, we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing paired fecal samples from mice that were collected at baseline and after clindamycin treatment. We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment in most mice (Fig. 3E and Table S9). When we compared the list of significant clindamycin impacted bacteria with the bacteria that varied between sources post-clindamycin, we found 4 OTUs that were shared between the lists (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23); Fig. 3D-E and Tables S8-9).  Importantly, some of the OTUs that varied between sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment (Fig. 3D), but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased in relative abundance after clindamycin treatment (Fig. 3D). These findings demonstrate that clindamycin had a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary between sources.

**Microbiota variation between sources is maintained after *C. difficile* challenge.** One day post-infection, significant differences in diversity metrics remained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Tables S1-2). Although the Charles River mice had more diverse communities and were also able to clear *C. difficile* faster than the other sources, diversity did not explain the observed variation in *C. difficile* colonization across sources. The Young and Schloss mice had the lowest diversity 1 day post-infection and were able to clear *C. difficile* earlier than Jackson, Taconic and Envigo mice. The source of mice and the interactions between source and cage effects continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`; *P* `r format.pval(p_adonis_all, eps = 0.001)`; Fig. 4C and Table S3). One day after *C. difficile* challenge, there were `r source_otus_d1` OTUs with significantly different relative abundances across sources (Fig. 4D and Table S10).

Throughout the experiment, the source of mice continued to be the dominant factor that explained the observed variation across fecal communities (PERMANOVA R^2^ = `r rsq_adonis_source`, *P* `r format.pval(p_adonis_source, eps = 0.001)`) followed by interactions between cage effects  and the day of the experiment (Movie S1 and Table S11). Fecal samples from the same source of mice continued to cluster closely to each other throughout the experiment. By 7 days post-infection, when approximately `r format(100-percent_colonized_d7, digits = 1)`% mice had cleared *C. difficile*, most of the mice had not recovered to their baseline community structure (Fig. 4E). The distance to the baseline community did not explain the variation in *C. difficile* clearance as the Schloss and Young mice had mostly cleared *C. difficile*, but their communities were a greater distance from baseline 7 days post-infection compared to the Jackson and Taconic mice that were still colonized. In summary, mouse bacterial communities varied significantly between sources throughout the course of the experiment and a consistent subset of bacteria remained different between sources regardless of clindamycin and *C. difficile* challenge.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
```
**Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C. difficile* by 7 days post-infection.** After identifying taxa that varied between sources, changed after clindamycin treatment, or both, we determined which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic regression models with either input bacterial community data from the baseline (day = -1), post-clindamycin (day = 0), or post-infection (day = 1) timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* `r format.pval(max_p_vs_random, eps = 0.001)`, Table S12). The model based on the post-clindamycin (AUROC = `r median_otu_day0_AUROC`) community OTU data performed significantly better than the baseline (AUROC = `r median_otu_dayn1_AUROC`) or the post-infection (AUROC = `r median_otu_day1_AUROC`) models (*P*~FDR~ `r format.pval(max_padj_otu_day0_comp, eps = 0.001)` for pairwise comparisons; Fig. S3C and Table S13). Thus, we were able to use bacterial relative abundance data from the time of *C. difficile* challenge to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile* at that timepoint. This result suggests that the bacterial community's response to clindamycin treatment had the greatest influence on subsequent *C. difficile* colonization dynamics.

To examine the bacteria that were driving each model’s performance, we selected the 20 OTUs that had the highest absolute feature weights in each of the 3 models (Table S14). First, we looked at OTUs from the model with the best performance, which was based on the post-clindamycin treatment (day 0) bacterial community data. Out of the 10 highest ranked OTUs, 7 OTUs were associated with *C. difficile* colonization 7 days post-infection (*Bacteroides*, *Escherichia/Shigella*, 2 *Lachnospiraceae*,  *Lactobacillus*, *Porphyromonadaceae*, and *Ruminococcaceae*), while 3 OTUs were associated with clearance (*Enterobacteriaceae*, *Lachnospiraceae*, *Porphyromonadaceae*; Fig. 5A). Next, we examined whether any of the top 20 ranked OTUs from the post-clindamycin (day 0) model were also important in the other 2 classification models based on baseline (day -1) and 1 day post-infection community data. We identified 6 OTUs that were important to the post-clindamycin model and either the baseline or 1 day post-infection models (*`r list_overlapping_across_models`*; Table S14). Thus, a subset of bacterial OTUs were important for determining *C. difficile* colonization dynamics across multiple timepoints.

To determine whether the OTUs driving the classification models also varied between sources, were altered by clindamycin treatment, or both, we identified the OTUs from each model that varied between sources (Fig. 1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features important to the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 5B). Together, these results suggest that the initial bacterial communities and their responses to clindamycin influenced the clearance of *C. difficile*.
